---
title: "Martinez-Ugalde et al 2025 Mexican soil bacterial atlas"
output: html_document
date: "Januray 2025"
---

Set working directory
```{r setup, include=FALSE}
setwd("~/Desktop/Atlas")
```

Loading R packages 
```{r Carga de paquetes, warning=FALSE, include=FALSE}
#General use and plot visualization
suppressMessages(library(qiime2R))
suppressMessages(library(tidyverse))
suppressMessages(library(readxl))
suppressMessages(library(paletteer))
suppressMessages(library(ggprism))
suppressMessages(library(ggplot2))
suppressMessages(library(ggpubr))
suppressMessages(library(patchwork))
suppressMessages(library(phyloseq))
suppressMessages(library(data.table))
suppressMessages(library(glue))
suppressMessages(library(ecole))
suppressMessages(library(vegan))
suppressMessages(library(ggcorrplot))
suppressMessages(library(ggrepel))
suppressMessages(library(ggdist))
suppressMessages(library(magrittr))
suppressMessages(library(ggnewscale))
suppressMessages(library(ggforce))
suppressMessages(library(ggalt))

#Libs for geographic data manipulation
suppressMessages(library(sf))
suppressMessages(library(geodata))
suppressMessages(library(terra))
suppressMessages(library(sp))
suppressMessages(library(microgeo))
suppressMessages(library(adespatial))

#Libs for map visualization
suppressMessages(library(leaflet))
suppressMessages(library(mapview))
suppressMessages(library(ggspatial))

#Libs for differential analysisis
suppressMessages(library(ANCOMBC))
suppressMessages(library(TreeSummarizedExperiment))
suppressMessages(library(ComplexHeatmap))
suppressMessages(library(circlize))
suppressMessages(library(colorspace))
```

Importing data to R
```{r Data loading, include=FALSE}
#Metadata
metadata = read.csv(file = "Metadata/Metadata_final_atlas.csv", 
                    sep = ",")

#Alpha div metrics
shannon = read_qza("Final_atlas_DS/Diversity_metrics/shannon_vector.qza")
obs_asv = read_qza("Final_atlas_DS/Diversity_metrics/observed_features_vector.qza")
faith = read_qza("Final_atlas_DS/Diversity_metrics/faith_pd_vector.qza")

#Beta div metrics
unifracMatx = read_qza("Final_atlas_DS/Diversity_metrics/weighted_unifrac_distance_matrix.qza")

#Phylogenetic tree
Phylo_tree = read_qza("Final_atlas_DS/Phylogeny/insertion-tree.qza")

#ASV table
ASV_Table = read_qza("Final_atlas_DS/Feature_table/Final_table_filtered_no_euk.qza")

#Representative sequecnes
Rep_seqs = read_qza("Final_atlas_DS/Rep_seqs/Final_seqs_filtered_no_euk.qza")

#Taxonomy
Taxonomy = read_qza("Final_set_atlas/All/All_taxonomy.qza",) 
```

Combine div metrics with metadata
```{r Data prep alpha and div metrics, include=FALSE}
shannon = shannon$data|>
  rownames_to_column() |>
  dplyr::rename(sampleID = rowname)

obs_asv = obs_asv$data |>
  rownames_to_column() |>
  dplyr::rename(sampleID = rowname)

faith = faith$data |>
  dplyr::rename(sampleID = V1)|>
  dplyr::rename(faithpd = V2)

metadata = metadata |>
  left_join(shannon)|>
  left_join(faith)|>
  left_join(obs_asv)

#Extract UniFrac matrix
weigh_mat = unifracMatx$data |> as.matrix()
```

Calculating alpha diversty statisitics
```{r Alpha diversity statistics, include=FALSE}
formulas = list(faithpd ~ Ecoregion,faithpd ~ Vegetation_cover,
                shannon_entropy ~ Ecoregion,shannon_entropy ~ Vegetation_cover,
                observed_features ~ Ecoregion,observed_features ~ Vegetation_cover)

#Krsukall test 
kruskal_test_results = function(formula) {
  result = kruskal.test(formula, data = metadata)
  data.frame(
    Formula = deparse(formula),
    Method = result$method,
    Chi_squared = result$statistic,
    Df = result$parameter,
    p_value = result$p.value
  )
}

kruskal_results = do.call(rbind, lapply(formulas, kruskal_test_results))

#Posthoc test
posthoc_wilcox_matrix = function(numeric_column, group_var) {
  pairwise_result = pairwise.wilcox.test(metadata[[numeric_column]], metadata[[group_var]],p.adjust.method = "BH")
  p_matrix = pairwise_result$p.value
  p_matrix[is.na(p_matrix)] = "NA"
  return(p_matrix)
}

posthoc_results_list = list()

for (formula in formulas) {
  kruskal_result = kruskal_test_results(formula)
  posthoc_results_list[[kruskal_result$Formula]] = list(Kruskal_Wallis = kruskal_result)
  response_var = all.vars(formula)[1] 
  group_var = all.vars(formula)[2]    
  
  if (kruskal_result$p_value < 0.05) {
    p_value_matrix = posthoc_wilcox_matrix(response_var, group_var)
    posthoc_results_list[[kruskal_result$Formula]]$Posthoc_Wilcox = p_value_matrix
    
    cat("\nPost-hoc Pairwise Wilcoxon Test p-value Matrix for",
        kruskal_result$Formula, ":\n")
    print(p_value_matrix)
  } else {
    cat("\nNo significant result from Kruskal-Wallis test for",
        kruskal_result$Formula, ", skipping post-hoc tests.\n")
  }
}
```

Create tables storing stat resuslts
```{r Table 1, include=FALSE}
Shannon_Ecoregion_2008_Conabio = posthoc_results_list[["shannon_entropy ~ Ecoregion"]]$Posthoc_Wilcox
Shannon_Ecoregion_2008_Conabio_KW = posthoc_results_list[["shannon_entropy ~ Ecoregion"]]$Kruskal_Wallis

Faith_Ecoregion_2008_Conabio = posthoc_results_list[["faithpd ~ Ecoregion"]]$Posthoc_Wilcox
Faith_Ecoregion_2008_Conabio_KW = posthoc_results_list[["faithpd ~ Ecoregion"]]$Kruskal_Wallis

Shannon_Vegetation_cover = posthoc_results_list[["shannon_entropy ~ Vegetation_cover"]]$Posthoc_Wilcox
Shannon_Vegetation_cover_KW = posthoc_results_list[["shannon_entropy ~ Vegetation_cover"]]$Kruskal_Wallis

Faith_Vegetation_cover = posthoc_results_list[["faithpd ~ Vegetation_cover"]]$Posthoc_Wilcox
Faith_Vegetation_cover_KW = posthoc_results_list[["faithpd ~ Vegetation_cover"]]$Kruskal_Wallis

results_kw = data.frame(Metric = c("Shannon Entropy", 
                                   "Faith's Phylogenetic Diversity"),
  Ecoregion_Chi2 = formatC(c(Shannon_Ecoregion_2008_Conabio_KW$Chi_squared,
                             Faith_Ecoregion_2008_Conabio_KW$Chi_squared),
                           digits = 4),
  Ecoregion_p = formatC(c(Shannon_Ecoregion_2008_Conabio_KW$p_value,
                          Faith_Ecoregion_2008_Conabio_KW$p_value),
                        digits = 4),
  Ecoregion_df = formatC(c(Shannon_Ecoregion_2008_Conabio_KW$Df,
                           Faith_Ecoregion_2008_Conabio_KW$Df),
                         digits = 4),
  Vegetation_cover_Chi2 = formatC(c(Shannon_Vegetation_cover_KW$Chi_squared,
                            Faith_Vegetation_cover_KW$Chi_squared),
                            digits = 4),
  Vegetation_cover_p = formatC(c(Shannon_Vegetation_cover_KW$p_value,
                         Faith_Vegetation_cover_KW$p_value),
                         digits = 4),
  Vegetation_cover_df = formatC(c(Shannon_Vegetation_cover_KW$Df,
                          Faith_Vegetation_cover_KW$Df+),
                          digits = 4),
  stringsAsFactors = FALSE)

write.csv(results_kw,"Final_atlas_DS/Diversity_stats/Kruskal_summ.csv", 
          row.names = FALSE)
```

Create a table with posthoc results
```{r Supplementary tables, include=FALSE}
combine_posthoc_matrices = function(shannon_matrix, faith_matrix, grouping_var) {
  
  shannon_df = as.data.frame(as.table(as.matrix(shannon_matrix)))
  faith_df = as.data.frame(as.table(as.matrix(faith_matrix)))

  colnames(shannon_df) = c("Group1", "Group2", "Shannon_p_value")
  colnames(faith_df) = c("Group1", "Group2", "Faith_p_value")
 
  
  combined_df = merge(shannon_df, faith_df, by = c("Group1", "Group2"), all = TRUE)
  return(combined_df)
}

shannon_ecoregion_matrix = posthoc_results_list[["shannon_entropy ~ Ecoregion"]]$Posthoc_Wilcox
faith_ecoregion_matrix = posthoc_results_list[["faithpd ~ Ecoregion"]]$Posthoc_Wilcox


ecoregion_combined = combine_posthoc_matrices(
  shannon_matrix = shannon_ecoregion_matrix, 
  faith_matrix = faith_ecoregion_matrix, 
  grouping_var = "Ecoregion"
)

shannon_Vegetation_cover_matrix = posthoc_results_list[["shannon_entropy ~ Vegetation_cover"]]$Posthoc_Wilcox
faith_Vegetation_cover_matrix = posthoc_results_list[["faithpd ~ Vegetation_cover"]]$Posthoc_Wilcox

Vegetation_cover_combined = combine_posthoc_matrices(
  shannon_matrix = shannon_Vegetation_cover_matrix, 
  faith_matrix = faith_Vegetation_cover_matrix, 
  grouping_var = "Land Use"
)

# Save results to CSV
write.csv(ecoregion_combined,
          "Final_atlas_DS/Diversity_stats/Ecoregion_Combined_Posthoc_f.csv", 
          row.names = FALSE)
write.csv(Vegetation_cover_combined,
          "Final_atlas_DS/Diversity_stats/Vegetation_cover_Combined_Posthoc_f.csv", 
          row.names = FALSE)
```

Beta diversity statistics
```{r PERMANOVA TEST, eval=FALSE, include=FALSE}
matrix_list = list(weigh_mat = weigh_mat,unweigh_mat = unweigh_mat,
                   jac_mat = jac_mat,bray_mat = bray_mat)  

variable_list = c("Ecoregion","Vegetation_cover")
variable_list_pair = c("Ecoregion","Vegetation_cover")
permut = 999

for (mat in names(matrix_list)) {
  for (var in variable_list) {
    formula_str = paste("matrix_list[['", mat, "']] ~ ", var, sep = "")
    formula = as.formula(formula_str)
    PERMANOVA_result = adonis2(formula,data = metadata, 
                                permutations = permut)
    df_name = paste("PERMANOVA", mat, var, sep = "_")
    assign(df_name, as.data.frame(PERMANOVA_result))
  }
}

Permanova_result_list = list(PERMANOVA_weigh_mat_Ecoregion,
                             PERMANOVA_weigh_mat_Vegetation_cover)

for (i in seq_along(Permanova_result_list)) {
  file_name = paste0("Permanova_result_", i, ".csv")
  write.csv(Permanova_result_list[[i]], file_name, row.names = FALSE)
}

PP_WUF_Ecosystem = permanova_pairwise(x=weigh_mat, grp = metadata$Ecoregion,
                                       padj = "BH",permutations = permut)

PP_WUF_Vegetation_cover = permanova_pairwise(x=weigh_mat, grp = metadata$Vegetation_cover,
                                       padj = "BH",permutations = permut)

write.csv(PP_WUF_Ecosystem, 
          "Final_atlas_DS/Diversity_stats/PP_WUF_Ecosystem.csv", row.names = FALSE)

write.csv(PP_WUF_Vegetation_cover, 
          "Final_atlas_DS/Diversity_stats/PP_WUF_Vegetation_cover.csv", row.names = FALSE)
```

PCoA calculation
```{r PCoA calculation, include=FALSE}
matrix_list = list(weigh_mat,unweigh_mat,jac_mat,bray_mat)  

PCoA_positions_list = list()
PCoA_percent_explained_list = list()

for (i in seq_along(matrix_list)) {
  matrix = matrix_list[[i]]
  pcoa_result = cmdscale(matrix, eig = TRUE, add = TRUE)
  positions = pcoa_result$points
  colnames(positions) = c("PCoA1", "PCoA2")
  percent_explained = 100 * pcoa_result$eig / sum(pcoa_result$eig)
  round_percentage = format(round(percent_explained[1:2], digits = 1), 
                            nsmall = 1, trim = TRUE)
  PCoA_positions_list[[i]] = positions
  PCoA_percent_explained_list[[i]] = round_percentage
  
  print(glue("Stored PCoA results for matrix {i}."))
}

weigh_PCOA_resutls = PCoA_positions_list[[1]]

weigh_PCOA_labs = PCoA_percent_explained_list[[1]]


lab_weigh = c(glue("PCoA 1 ({weigh_PCOA_labs[1]}%)"),
              glue("PCoA 2 ({weigh_PCOA_labs[2]}%)"))
```

RDA with variance partitioning
```{r RDA with var part}
#Response matrix
atlas_dist_weig_hell = decostand(weigh_mat, "hellinger")

#Explinatory datastes
clim_var = metadata[,c(1,38,39:57,58)]
edap_var = metadata[,c(1,30:37)]

eco_var = metadata[,c(1,26)]
luse_var = metadata[,c(1,13)]

#Data preparation
clim_scale = c("Bio1","Bio2","Bio3",
               "Bio4","Bio5","Bio6",
               "Bio7","Bio8","Bio9",
               "Bio10","Bio11","Bio12",
               "Bio13","Bio14","Bio15",
               "Bio16","Bio17","Bio18",
               "Bio19","AI","ELEV")

edap_scale = c("bdod","cfvo","nitrogen","phh2o","sand","silt","soc","ocd")

eco_cat = c("Ecoregion")
luse_cat = c("Vegetation_cover")

#Order data in the same way as distance matrix
clim_var = clim_var |>
  dplyr::select(sampleID, all_of(clim_scale)) |>
  column_to_rownames(var = "sampleID") |>
  as.data.frame()|>
  rownames_to_column(var = "sampleID") |>
  dplyr::arrange(match(sampleID, rownames(weigh_mat))) |>
  column_to_rownames(var = "sampleID")

edap_var = edap_var |>
  dplyr::select(sampleID, all_of(edap_scale)) |>
  column_to_rownames(var = "sampleID") |>
  as.data.frame()|>
  rownames_to_column(var = "sampleID") |>
  dplyr::arrange(match(sampleID, rownames(weigh_mat))) |>
  column_to_rownames(var = "sampleID")

eco_var = eco_var |>
  dplyr::select(sampleID, all_of(eco_cat)) |>
  column_to_rownames(var = "sampleID") |>
  as.data.frame()|>
  rownames_to_column(var = "sampleID") |>
  dplyr::arrange(match(sampleID, rownames(weigh_mat))) |>
  column_to_rownames(var = "sampleID")

luse_var = luse_var |>
  dplyr::select(sampleID, all_of(luse_cat)) |>
  column_to_rownames(var = "sampleID") |>
  as.data.frame()|>
  rownames_to_column(var = "sampleID") |>
  dplyr::arrange(match(sampleID, rownames(weigh_mat))) |>
  column_to_rownames(var = "sampleID")

#Scaling data and converting dummy vars
clim_var[clim_scale] = scale(clim_var[clim_scale])
edap_var[edap_scale] = scale(edap_var[edap_scale])

eco_var$Ecoregion = as.factor(eco_var$Ecoregion)
luse_var$Vegetation_cover = as.factor(luse_var$Vegetation_cover)
```

Run ordistep selection for climatic, edaphic variables
```{r Forward selection (clim / edaph), eval=FALSE, include=FALSE}
dbrda_climatic = capscale(atlas_dist_weig_hell ~ ., clim_var)
dbrda_climatic_0 = capscale(atlas_dist_weig_hell ~ 1, clim_var)

dbrda_edaphic = capscale(atlas_dist_weig_hell ~ ., edap_var)
dbrda_edaphic_0 = capscale(atlas_dist_weig_hell ~ 1, edap_var)

clim_for_sel = ordiR2step(dbrda_climatic_0,
                          scope = formula(dbrda_climatic),
                          direction = "both",
                          R2scope = TRUE)

edap_for_sel = ordiR2step(dbrda_edaphic_0,
                          scope = formula(dbrda_edaphic),
                          direction = "both",
                          R2scope = TRUE)

selected_climatic = clim_var[, colnames(clim_var) %in%
                               attr(terms(clim_for_sel), "term.labels")]

selected_edaph = edap_var[, colnames(edap_var) %in%
                               attr(terms(edap_for_sel), "term.labels")]

#Removing iterativley variables with high vif
clim_vif_values = vif.cca(clim_for_sel)
edap_vif_values = vif.cca(edap_for_sel)

while (any(clim_vif_values > 10)) {
  high_vif_var_c = names(which.max(clim_vif_values))
  updated_formula_c = update(clim_for_sel$call$formula, 
                            paste(". ~ . -", high_vif_var_c))
  clim_for_sel = capscale(updated_formula_c, data = clim_var)
  clim_vif_values = vif.cca(clim_for_sel)
}

while (any(edap_vif_values > 10)) {
  high_vif_var_e = names(which.max(edap_vif_values))
  updated_formula_e = update(edap_for_sel$call$formula, 
                            paste(". ~ . -", high_vif_var_e))
  edap_for_sel = capscale(updated_formula_e, data = edap_var)
  edap_vif_values = vif.cca(edap_for_sel)
}

#Selected vars
print(clim_for_sel$call)
clim_selected = clim_var |>
  dplyr::select(Bio19,Bio14,
                Bio3,Bio8,
                Bio15,Bio2,
                Bio18)|>
  rownames_to_column(var = "sampleID")

print(edap_for_sel$call)
edap_selected = edap_var |>
  dplyr::select(phh2o,soc,
                nitrogen,cfvo,
                bdod,sand,silt)|>
  rownames_to_column(var = "sampleID")

#Creating a final df
explanatory_vars = clim_selected|>
  left_join(edap_selected)

eco_var = eco_var |>
  rownames_to_column(var = "sampleID")
luse_var = luse_var |>
  rownames_to_column(var = "sampleID")

covariates_vars = eco_var|>
  left_join(luse_var)

dbrda_data_f = explanatory_vars|>
  left_join(covariates_vars)|>
  column_to_rownames(var = "sampleID")
```

Run partial dbRDA
```{r Perform RDA}
dbrda_data_f$Ecoregion = as.factor(dbrda_data_f$Ecoregion)
dbrda_data_f$Vegetation_cover = as.factor(dbrda_data_f$Vegetation_cover)

db_RDA_sol_atlas = capscale(atlas_dist_weig_hell ~ phh2o + soc + nitrogen + cfvo +
                           bdod + sand + silt + Bio19 + Bio3 + Bio2 + Bio8 + Bio15 + 
                          Bio14 + Bio14 + Bio18 + 
                            Condition (Ecoregion + Vegetation_cover),
                         data = dbrda_data_f)

db_rda_baseplot = plot(db_RDA_sol_atlas)
```

Runs statistics test for dbRDA
```{r RDA statistics}
#Analyse RDA resutls
smry

#PERFORM ANOVA for terms
RDA_res_var = anova.cca(db_RDA_sol_atlas,by = "terms")

#PERFORM ANOVA for margins
RDA_res_mar = anova.cca(db_RDA_sol_atlas,by = "margin")

#PERFORM ANOVA for axis
RDA_res_axis = anova.cca(db_RDA_sol_atlas,by= "axis")

#Recover coefficients
RDA_res_coef = coef(db_RDA_sol_atlas)

#Recover R2
RDA_res_r2 = RsquareAdj(db_RDA_sol_atlas)
```

Microgeo
```{r Data gathering for microgeo, include=FALSE}
#Map obtention
mex_map = read_shp_map("Mapas/estadosmx/estadosmx.shp")
mex_map_microgeo = trans_map_fmt(mex_map, var = "nom_ent")

#Data edition
Taxonomy_microgeo = Taxonomy_ancom |>
  dplyr::select(-Class_S)|>
  mutate(Domain = ifelse(is.na(Domain), "d__", Domain))|>
  mutate(Phyla = ifelse(is.na(Phyla), "p__", Phyla))|>
  mutate(Class = ifelse(is.na(Class), "c__", Class))|>
  mutate(Family = ifelse(is.na(Family), "f__", Family))|>
  mutate(Genus = ifelse(is.na(Genus), "g__", Genus))|>
  column_to_rownames(var = "Feature.ID")
  
Taxonomy_microgeo = Taxonomy_microgeo[match(rownames(ASV_Table_microgeo),
                                      rownames(Taxonomy_microgeo)),]

metadata_microgeo = metadata_microgeo |>
  dplyr::select(sampleID,Latitude,Longitude)|>
  column_to_rownames(var = "sampleID")


Phylo_tree = Phylo_tree$data 

atlas_microgeo_data = create_dataset(map = mex_map_microgeo,
                                     mat = ASV_Table_microgeo,
                                     ant = Taxonomy_microgeo,
                                     met = metadata_microgeo,
                                     phy = Phylo_tree,
                                     lon = "Longitude",
                                     lat = "Latitude")

atlas_microgeo_data %<>% get_soilgrid(measures = c("bdod", "cfvo","nitrogen",
                                                   "phh2o", "sand", "silt",
                                                   "soc", "ocd"),depth = 15)

atlas_microgeo_data %<>% get_ai(out.dir = "test_f") 
atlas_microgeo_data %<>% get_his_bioc(res = 0.5, out.dir = "test_f")
atlas_microgeo_data %<>% get_elev(res = 0.5, out.dir = "test_f")
atlas_microgeo_data %<>% extract_data_from_spatraster(type = "his")
atlas_microgeo_data %<>% get_modis_cla_metrics(username = "martugal", 
                                               password = "Machamp50@@@",
                                               measures = c("LC_Type1","LC_Type2",
                                                            "LC_Type3"),
                                               out.dir = "test_f")

atlas_microgeo_data %<>% rarefy_count_table(depth = 40000)
atlas_microgeo_data %<>% tidy_dataset()
atlas_microgeo_data %<>% calc_rel_abund()
```

Recover metadata gathered from microgeo
```{r Recover metadta from microgeo, eval=FALSE, include=FALSE}
microgeo_meta = atlas_microgeo_data$spa$tabs
microgeo_meta = microgeo_meta |>
  rownames_to_column()|>
  dplyr::rename(sampleID = rowname)
write_csv(microgeo_meta,"Final_atlas_DS/Metadata/microgeo_meta.csv")
```

Calculate biogeographical traits 
```{r Calculating ecological markers, include=FALSE}
ph_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "phh2o", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

soc_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "soc", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

nit_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "nitrogen", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

cfvo_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "cfvo", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

bdod_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "bdod", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

sand_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "sand", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

silt_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "silt", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

bio19_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "Bio19", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

bio14_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "Bio14", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

bio3_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "Bio3", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

bio8_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "Bio8", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

bio15_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "Bio15", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

bio2_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "Bio2", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)

bio18_assocated_traits = atlas_microgeo_data %<>% 
  calc_markers(use.var = "Bio18", annotation.level = "Order", 
               r.thres = 0.4, use.dat = "spa",
               p.adjust.method = "BH",p.thres = 0.001)
```

Verify microgeo results
```{r Storing ecological marker results in tables, include=FALSE}
ph_makers = ph_assocated_traits$abd$mar$correlation
soc_makers = soc_assocated_traits$abd$mar$correlation
nit_makers = nit_assocated_traits$abd$mar$correlation
cfvo_makers = cfvo_assocated_traits$abd$mar$correlation
bdod_makers = bdod_assocated_traits$abd$mar$correlation
silt_makers = silt_assocated_traits$abd$mar$correlation
bio19_makers = bio19_assocated_traits$abd$mar$correlation
bio8_makers = bio8_assocated_traits$abd$mar$correlation
bio2_makers = bio2_assocated_traits$abd$mar$correlation
bio3_makers = bio3_assocated_traits$abd$mar$correlation
bio18_makers = bio18_assocated_traits$abd$mar$correlation
bio14_makers = bio14_assocated_traits$abd$mar$correlation
###
ph_makers = ph_makers |>
  mutate(Category = "ph")

soc_makers = soc_makers |>
  mutate(Category = "soc")

nit_makers = nit_makers|>
  mutate(Category = "nit")

cfvo_makers = cfvo_makers|>
  mutate(Category = "cfvo")

bdod_makers = bdod_makers|>
  mutate(Category = "bdod")

silt_makers = silt_makers|>
  mutate(Category = "silt")

bio19_makers = bio19_makers|>
  mutate(Category = "bio19")

bio8_makers = bio8_makers|>
  mutate(Category = "bio8")

bio2_makers = bio2_makers|>
  mutate(Category = "bio2")

bio3_makers = bio3_makers|>
  mutate(Category = "bio3")

bio18_makers = bio18_makers|>
  mutate(Category = "bio18")

bio14_makers = bio14_makers|>
  mutate(Category = "bio14")
###
marker_list = list(ph_makers,soc_makers,nit_makers,cfvo_makers, 
               silt_makers,bio19_makers,bio8_makers,bio2_makers,
               bio3_makers,bio18_makers,bio14_makers)

merged_mark = purrr::reduce(marker_list, full_join, by = c("var"))

category_cols = names(merged_mark)[grepl("^Category", names(merged_mark))]

merged_mark = merged_mark |>
  rowwise() |>
  mutate(Category = paste(unique(na.omit(c_across(all_of(category_cols)))), collapse = ", ")) |>
  ungroup() |>
  select(var, Category)|>
  dplyr::rename(Order = var)

Taxonomy_ancom_g = Taxonomy_ancom |>
  distinct(Order, .keep_all = TRUE) |>
  select(-Feature.ID)

merged_mark_f = merged_mark |>
  left_join(Taxonomy_ancom_g)

merged_mark_f |> 
  count(value = "Phyla")

write_csv(merged_mark_f, "Final_atlas_DS/Diversity_stats/Markers.csv")

```

Evaluate if predicted markers are present at high or low abudances
```{r Classify ecological markeres, include=FALSE}
#If the relative abundance is more than 5%, in 50 orders were classified with high abudn

#Obtain the list of marker taxa
ph_list = rownames(ph_makers) |> as.character()
soc_list = rownames(soc_makers) |> as.character()
nit_list = rownames(nit_makers) |> as.character()
cfvo_list = rownames(cfvo_makers) |> as.character()
bdod_list = rownames(bdod_makers) |> as.character()
silt_list = rownames(silt_makers) |> as.character()
bio8_list = rownames(bio8_makers) |> as.character()
bio19_list = rownames(bio19_makers) |> as.character()
bio2_list = rownames(bio2_makers) |> as.character()
bio3_list = rownames(bio3_makers) |> as.character()
bio18_list = rownames(bio18_makers) |> as.character()
bio14_list = rownames(bio14_makers) |> as.character()

# Function to process each taxon
process_taxa = function(taxon) {
  df = data.frame(
    row.names = rownames(atlas_microgeo_data$abd$raw$Order),
    Taxon = taxon,
    Abundance = atlas_microgeo_data$abd$raw$Order[[taxon]]
  )
  
  df$Abundance_Class = ifelse(df$Abundance > 1, "Prevalent", "No prevalent")
  
  df$Abundance_Class = as.factor(df$Abundance_Class)
  
  return(df)
}

# Apply function to all taxa and combine into a single table
ph_final_table = do.call(rbind, lapply(ph_list, process_taxa))
soc_final_table = do.call(rbind, lapply(soc_list, process_taxa))
nit_final_table = do.call(rbind, lapply(nit_list, process_taxa))
cfvo_final_table = do.call(rbind, lapply(cfvo_list, process_taxa))
bdod_final_table = do.call(rbind, lapply(bdod_list, process_taxa))
silt_final_table = do.call(rbind, lapply(silt_list, process_taxa))
bio8_final_table = do.call(rbind, lapply(bio8_list, process_taxa))
bio19_final_table = do.call(rbind, lapply(bio19_list, process_taxa))
bio2_final_table = do.call(rbind, lapply(bio2_list, process_taxa))
bio3_final_table = do.call(rbind, lapply(bio3_list, process_taxa))
bio18_final_table = do.call(rbind, lapply(bio18_list, process_taxa))
bio14_final_table = do.call(rbind, lapply(bio14_list, process_taxa))

# Print summary table
ph_clas_table = table(ph_final_table$Taxon, ph_final_table$Abundance_Class)
soc_clas_table = table(soc_final_table$Taxon, soc_final_table$Abundance_Class)
nit_clas_table = table(nit_final_table$Taxon, nit_final_table$Abundance_Class)
cfvo_clas_table = table(cfvo_final_table$Taxon, cfvo_final_table$Abundance_Class)
bdod_clas_table = table(bdod_final_table$Taxon, bdod_final_table$Abundance_Class)
silt_clas_table = table(silt_final_table$Taxon, silt_final_table$Abundance_Class)
bio8_clas_table = table(bio8_final_table$Taxon, bio8_final_table$Abundance_Class)
bio19_clas_table = table(bio19_final_table$Taxon, bio19_final_table$Abundance_Class)
bio2_clas_table = table(bio2_final_table$Taxon, bio2_final_table$Abundance_Class)
bio3_clas_table = table(bio3_final_table$Taxon, bio3_final_table$Abundance_Class)
bio18_clas_table = table(bio18_final_table$Taxon, bio18_final_table$Abundance_Class)
bio14_clas_table = table(bio14_final_table$Taxon, bio14_final_table$Abundance_Class)
####
ph_prevalent = ph_clas_table |> as.data.frame() |> 
  dplyr::filter(Var2 == "Prevalent" & Freq > 251) |>
  mutate(Category = "pH")

soc_prevalent = soc_clas_table |> as.data.frame() |> 
  dplyr::filter(Var2 == "Prevalent" & Freq > 251) |>
  mutate(Category = "soc")

nit_prevalent = nit_clas_table |> as.data.frame() |> 
  dplyr::filter(Var2 == "Prevalent" & Freq > 251) |>
  mutate(Category = "nit")

bdod_prevalent = bdod_clas_table |> as.data.frame() |> 
  dplyr::filter(Var2 == "Prevalent" & Freq > 251) |>
  mutate(Category = "bdod")

silt_prevalent = silt_clas_table |> as.data.frame() |> 
  dplyr::filter(Var2 == "Prevalent" & Freq > 251) |>
  mutate(Category = "silt")

bio8_prevalent = bio8_clas_table |> as.data.frame() |> 
  dplyr::filter(Var2 == "Prevalent" & Freq > 251) |>
  mutate(Category = "bio8")

bio19_prevalent = bio19_clas_table |> as.data.frame() |> 
  dplyr::filter(Var2 == "Prevalent" & Freq > 251) |>
  mutate(Category = "bio19")

bio2_prevalent = bio2_clas_table |> as.data.frame() |> 
  dplyr::filter(Var2 == "Prevalent" & Freq > 251) |>
  mutate(Category = "bio2")

bio3_prevalent = bio3_clas_table |> as.data.frame() |> 
  dplyr::filter(Var2 == "Prevalent" & Freq > 251) |>
  mutate(Category = "bio3")

bio18_prevalent = bio18_clas_table |> as.data.frame() |> 
  dplyr::filter(Var2 == "Prevalent" & Freq > 251) |>
  mutate(Category = "bio18")

bio14_prevalent = bio14_clas_table |> as.data.frame() |> 
  dplyr::filter(Var2 == "Prevalent" & Freq > 251) |>
  mutate(Category = "bio14")
###
df_list = list(ph_prevalent,soc_prevalent,nit_prevalent,bdod_prevalent, 
               bio8_prevalent,bio19_prevalent,bio2_prevalent,silt_prevalent,
               bio3_prevalent,bio18_prevalent,bio14_prevalent)

merged_df = purrr::reduce(df_list, full_join, by = c("Var1", "Freq"))

category_cols = names(merged_df)[grepl("^Category", names(merged_df))]

merged_df = merged_df |>
  rowwise() |>
  mutate(Category = paste(unique(na.omit(c_across(all_of(category_cols)))), collapse = ", ")) |>
  ungroup() |>
  select(Var1, Freq, Category)
```

Z-scoring abudance table 
```{r Normilize abd table}
zscore_table = atlas_microgeo_data$abd$raw$Order
zscore_table = scale(zscore_table) |>
  as.data.frame()
```

Ecological marker maps
```{r Creating maps}
##
Subgroup_2_rf_mod = create_ml_model(y.data = zscore_table, 
                           x.data = atlas_microgeo_data$spa$tabs,
                           var = "o__Subgroup_2", method = "rf")

Subgroup_2_rf_geo = Subgroup_2_rf_mod |>
  predict_ml_geomap(spat.raster = atlas_microgeo_data$spa$rast$his)

Subgroup_2_map = plot_bmap(map = atlas_microgeo_data$map) |>
  add_spatraster(spat.raster = Subgroup_2_rf_geo,
                 legend.title = "Subgroup_2\n PRA") |>
  add_scale_bar() |>
  add_north_arrow()

Subgroup_2_map = Subgroup_2_map +
  labs(title = "A")
##
Acidobacteriales_rf_mod = create_ml_model(y.data = zscore_table, 
                           x.data = atlas_microgeo_data$spa$tabs,
                           var = "o__Acidobacteriales", method = "rf")

Acidobacteriales_rf_geo = Acidobacteriales_rf_mod |>
  predict_ml_geomap(spat.raster = atlas_microgeo_data$spa$rast$his)

Acidobacteriales_map = plot_bmap(map = atlas_microgeo_data$map) |>
  add_spatraster(spat.raster = Acidobacteriales_rf_geo,
                 legend.title = "Acidobacteriales\n PRA") |>
  add_scale_bar() |>
  add_north_arrow()

Acidobacteriales_map = Acidobacteriales_map +
  labs(title = "B")
##
Solibacterales_rf_mod = create_ml_model(y.data = zscore_table, 
                           x.data = atlas_microgeo_data$spa$tabs,
                           var = "o__Solibacterales", method = "rf")

Solibacterales_rf_geo = Solibacterales_rf_mod |>
  predict_ml_geomap(spat.raster = atlas_microgeo_data$spa$rast$his)

Solibacterales_map = plot_bmap(map = atlas_microgeo_data$map) |>
  add_spatraster(spat.raster = Solibacterales_rf_geo,
                 legend.title = "Solibacterales\n PRA") |>
  add_scale_bar() |>
  add_north_arrow()

Solibacterales_map = Solibacterales_map +
  labs(title = "C")
##
Alphaproteobacteria_rf_mod = create_ml_model(y.data = zscore_table, 
                           x.data = atlas_microgeo_data$spa$tabs,
                           var = "o__Alphaproteobacteria", method = "rf")

Alphaproteobacteria_rf_geo = Alphaproteobacteria_rf_mod |>
  predict_ml_geomap(spat.raster = atlas_microgeo_data$spa$rast$his)

Alphaproteobacteria_map = plot_bmap(map = atlas_microgeo_data$map) |>
  add_spatraster(spat.raster = Alphaproteobacteria_rf_geo,
                 legend.title = "Unclassified Alphaproteobacteria\n PRA") |>
  add_scale_bar() |>
  add_north_arrow()

Alphaproteobacteria_map = Alphaproteobacteria_map +
  labs(title = "D")
##
Rhizobiales_rf_mod = create_ml_model(y.data = zscore_table, 
                           x.data = atlas_microgeo_data$spa$tabs,
                           var = "o__Rhizobiales", method = "rf")

Rhizobiales_rf_geo = Rhizobiales_rf_mod |>
  predict_ml_geomap(spat.raster = atlas_microgeo_data$spa$rast$his)

Rhizobiales_map = plot_bmap(map = atlas_microgeo_data$map) |>
  add_spatraster(spat.raster = Rhizobiales_rf_geo,
                 legend.title = "Rhizobiales\n PRA") |>
  add_scale_bar() |>
  add_north_arrow()

Rhizobiales_map = Rhizobiales_map +
  labs(title = "E")
##
Chthoniobacterales_rf_mod = create_ml_model(y.data = zscore_table, 
                           x.data = atlas_microgeo_data$spa$tabs,
                           var = "o__Chthoniobacterales", method = "rf")

Chthoniobacterales_rf_geo = Chthoniobacterales_rf_mod |>
  predict_ml_geomap(spat.raster = atlas_microgeo_data$spa$rast$his)

Chthoniobacterales_map = plot_bmap(map = atlas_microgeo_data$map) |>
  add_spatraster(spat.raster = Chthoniobacterales_rf_geo,
                 legend.title = "Chthoniobacterales\n PRA") |>
  add_scale_bar() |>
  add_north_arrow()

Chthoniobacterales_map = Chthoniobacterales_map +
  labs(title = "F")
##
Gemmatimonadales_rf_mod = create_ml_model(y.data = zscore_table, 
                           x.data = atlas_microgeo_data$spa$tabs,
                           var = "o__Gemmatimonadales", method = "rf")

Gemmatimonadales_rf_geo = Gemmatimonadales_rf_mod |>
  predict_ml_geomap(spat.raster = atlas_microgeo_data$spa$rast$his)

Gemmatimonadales_map = plot_bmap(map = atlas_microgeo_data$map) |>
  add_spatraster(spat.raster = Gemmatimonadales_rf_geo,
                 legend.title = "Gemmatimonadales\n PRA") |>
  add_scale_bar() |>
  add_north_arrow()

Gemmatimonadales_map = Gemmatimonadales_map +
  labs(title = "G")
##
Vicinamibacterales_rf_mod = create_ml_model(y.data = zscore_table, 
                           x.data = atlas_microgeo_data$spa$tabs,
                           var = "o__Vicinamibacterales", method = "rf")

Vicinamibacterales_rf_geo = Vicinamibacterales_rf_mod |>
  predict_ml_geomap(spat.raster = atlas_microgeo_data$spa$rast$his)

Vicinamibacterales_map = plot_bmap(map = atlas_microgeo_data$map) |>
  add_spatraster(spat.raster = Vicinamibacterales_rf_geo,
                 legend.title = "Vicinamibacterales\n PRA") |>
  add_scale_bar() |>
  add_north_arrow()

Vicinamibacterales_map = Vicinamibacterales_map +
  labs(title = "H")
```